Towards Generic Parallel Programming in Computer Science Education with Kokkos

### 2020-08-15 src/0.Abstract.tex
Updates for submission to EduHPC at SC20

\begin{abstract}\label{chap:abstract}
Parallel patterns, views, and spaces are promising abstractions to capture the programmer's intent as well as the contextual information that can be used by an underlying runtime to efficiently map software to parallel hardware. These abstractions can be valuable in cases where an algorithm must accommodate requirements of code and performance portability across hardware architectures and vendor programming models. Kokkos is a parallel programming model for host- and accelerator architectures that relies on these abstractions and targets these requirements. It consists of a pure C++ interface, a specification, and a programming library. The programming library exposes patterns and types and maps them to an underlying abstract machine model. The abstract machine model offers a generic view of parallel hardware. While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to parallel programming productivity. In this work, we give an insight into design considerations of a representative programming model, list important abstractions, document feedback received during workshops, and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
\end{abstract}

>>>>

\begin{abstract}\label{chap:abstract}
Parallel patterns, views, and spaces are promising abstractions to capture the programmer's intent as well as the contextual information that can be used by an underlying runtime to efficiently map software to parallel hardware. These abstractions can be valuable in cases where an algorithm must accommodate requirements of code and performance portability across hardware architectures and vendor programming models. Kokkos is a parallel programming model for host- and accelerator architectures that relies on these abstractions and targets these requirements. It consists of a pure C++ interface, a specification, and a programming library. The programming library exposes patterns and types and maps them to an underlying abstract machine model. The abstract machine model offers a generic view of parallel hardware. While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to a generic, vendor, and architecture-independent education of parallel programming. In this work, we give an insight into the design considerations of this programming model and list important abstractions. Further, we document best practices obtained from giving virtual classes on Kokkos, document attendee feedback, and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
\end{abstract}

### 2020-08-15 src/1.Introduction.tex
Updates for submission to EduHPC at SC20

Parallel and distributed computing is a focus at universities and teaching institutions. The interest is two-fold. Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations.

While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined. In this case, educators can select from a variety of solutions, approaches, and trends in parallel computing. Which particular set to choose for teaching in class is not trivial. Complexity is further amplified by the breath of related topics ranging from parallel hardware architectures, programming models, algorithms as well as productivity in parallel software development. However, it is the productivity concern that is gaining importance as the number of hardware solutions, their complexity as well as demands for a quick time-to-solution increase.

While vendor programming models maintain their position in curricula today, we think that the inclusion of parallel programming models such as Kokkos is worth consideration. They offer an easy to understand set of concepts for a vendor- and architecture-independent, performance-portable expression of concurrency which significantly contributes to parallel programming productivity.

@@

Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions in parallel programming models. We show a set of interfaces that implement patterns, views, and spaces, and capture enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++. For teaching insights, we document feedback received during on-line lectures.

The rest of this paper is structured as follows. The next chapter gives an overview of parallel programming concepts and presents the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive the required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application. In Chapter~\ref{chap:backend} we provide insight into a Kokkos back-end implementation. Chapters~\ref{chap:related} and ~\ref{chap:tutorials} discuss related work, tutorials and student feedback from tutorials. Lastly Chapter~\ref{chap:conclusion} concludes this work.

>>>>

Parallel computing is an area of academic focus. The interest is two-fold. Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations.

While industry-driven collaborations often predefine technological frameworks such as programming environments and target architectures, academic teaching leaves this mostly undefined. In this case, educators can select from a variety of architectures and programming models for their parallel computing curricula. Which particular set to choose for teaching in class is not trivial. While decision factors differ, maximizing the subject relevance by teaching generic principles of parallel programming is among the most important ones.

Consequently, generic parallel programming models are desirable that target mainstream programming languages and allow a vendor- and architecture-independent, performance-portable expression of concurrency. In this way, a parallel program becomes capable of running efficiently on computer architectures that conform to a common parallel machine specification. We would like to compare this to the support of Von Neumann architectures in sequential programming languages today.

While message-passing interfaces, OpenMP, OpenCL or CUDA maintain their role in teaching, the plethora of novel parallel programming models~\cite{Chapel}~\cite{UPC++}~\cite{Raja} hints towards an emphasis on productivity in parallel programming and also brings up the valid question whether this strengthens the argument for their inclusion in education.

Based on our experience in supercomputing, we see that productivity is gaining importance as the number of hardware solutions, their complexity to program as well as demands for a quick time-to-solution increase. Consequently, we think that the inclusion of novel programming models is worth consideration in case they offer generic programming concepts for a vendor- and architecture-independent, performance-portable expression of concurrency and can contribute to parallel programming productivity.

It is the productivity concern as well as the need for a generic expression of concurrency due to periodic changes of supercomputer architectures and vendors in national projects that inspired the development of Kokkos. The increasing adoption of Kokkos in these projects confirms its usefulness which motivates us to present Kokkos to academia in this work.

@@

Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions in parallel programming models. We show a set of interfaces that implement patterns, views, and spaces, and capture enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++. For teaching insights, we document our best practices obtained from teaching Kokkos as well as feedback received during on-line lectures.

The rest of this paper is structured as follows. The next chapter gives an overview of parallel programming concepts and presents the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive the required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application. In Chapter~\ref{chap:backend} we provide insight into a Kokkos backend implementation. Chapters~\ref{chap:related} and ~\ref{chap:teaching} discuss related work, best practices in teaching, and student feedback from tutorials. Lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-08-15 src/2.GenericProgramming.tex
Updates for submission to EduHPC at SC20

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance-portable code. This is commonly achieved by adding annotations to the base programming language, often through \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with a sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

@@

A \emph{Parallel Paradigm} is an abstract representation to facilitate the programmer's understanding of programming rules and program behavior. Which programming paradigm to choose depends on several considerations. Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that supports the expression of interdependent work items needed for irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels. This paradigm eliminates access to a shared state and aligns well with message passing programming (MPI). The execution model and memory model aspects of a programming model define its behavior. That is, it defines the relationship between abstract concepts and program execution on the given architecture.

>>>>

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to add concurrency to their code. This is commonly achieved by adding annotations to the base programming language, often through \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with a sufficiently constrained semantics that is used to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

@@

A \emph{Parallel Paradigm} is an abstract representation to facilitate the programmer's understanding of programming rules and program behavior. Which programming paradigm to choose depends on several considerations. Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that supports the expression of interdependent work items needed for irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels. This paradigm eliminates access to a shared state and aligns well with message passing programming (MPI). The execution model and memory model aspects of a programming model define its behavior. That is, it defines the relationship between abstract concepts and program execution on a given hardware architecture.

### 2020-08-15 src/3.MachineModel.tex
Updates for submission to EduHPC at SC20

The Kokkos machine model defines abstractions that represent hardware capabilities for processing and data access. The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API. By abstracting from physical hardware, the machine model ensures that applications written in programming models targeting this machine model are generic and performance-portable on current and future hardware. The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model. We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristics and portability remain the same.

@@

\includegraphics[width=0.5\textwidth]{img/KokkosSpaces.png}
\caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
\label{fig:execspace}
\label{fig:spaces}
\end{figure}

>>>>

The Kokkos machine model defines abstractions that represent hardware capabilities for processing and data access. The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API. By abstracting from physical hardware, the machine model ensures that applications written in any programming model that is targeting this machine model are generic and performance-portable on current and future hardware. Kokkos is one particular instance of a programming model that builds on top of that machine model. We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristics and portability remain the same.

@@

\includegraphics[width=0.49\textwidth]{img/KokkosSpaces.png}
\centering
\caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
\label{fig:execspace}
\label{fig:spaces}
\end{figure}

### 2020-08-15 src/4.ProgammingModel.tex
Updates for submission to EduHPC at SC20

The Kokkos programming model specifies a programming paradigm and an API. The API exposes the machine representation to the developer and defines a set of patterns and types that capture intents and properties of execution. This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.

Kokkos defines a pure interface for C++ and uses C++ as a programming language. This has multiple reasons. It is our experience that application developers of scientific and high-performance computing applications in the high-performance computing community and in the industry in general express the demand for C++ and a generic parallel programming model to support their codes. Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represents a burden worth addressing. Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.

@@

Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought. Kokkos supports parallel patterns and tasking. Parallel patterns covers \emph{for}, \emph{scan} and \emph{reduction}. This allows the expression of concurrency over iterative, for-loop computable algorithms. To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm. Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program. Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and capture the intent of the programmer.

@@

To the interested reader, the complete programming model description can be accessed on-line\cite{KOKKOS_WIKI}.
The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the underlying back-end programming model.

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{img/Abstractions.png}}
\caption{Overview of abstractions that define the constrained semantic of the Kokkos parallel programming model.}
\label{fig:abstractions}
\end{figure}

>>>>

The Kokkos programming model specifies a programming paradigm and an API. The API exposes the machine representation to the developer and defines a set of patterns and types that capture intents and properties of execution. This completes the semantic capture, discussed in Chapter~\ref{chap:background}. We detail all three components as follows.

Kokkos defines a pure interface for C++ and uses C++ as a programming language. This has multiple reasons. It is our experience that developers of scientific and high-performance computing applications express the demand for C++ and a generic parallel programming model to support their codes. Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represents a burden worth addressing. Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.

@@

Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought. Kokkos supports parallel patterns and tasking. Parallel patterns cover \emph{for}, \emph{scan} and \emph{reduction}. This allows the expression of concurrency over iterative, for-loop computable algorithms. To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm. Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program. Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and capture the intent of the programmer.

@@

To the interested reader, the complete programming model description can be accessed on-line\cite{KOKKOS_WIKI}.
The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the underlying backend programming model.

\begin{figure}
\centerline{\includegraphics[width=0.49\textwidth]{img/Abstractions.png}}
\caption{Overview of abstractions that define the constrained semantic of the Kokkos parallel programming model.}
\label{fig:abstractions}
\end{figure}

### 2020-08-15 src/5.Example.tex
Updates for submission to EduHPC at SC20

If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation. In real applications, a host-accessible allocation is necessary to allow I/O operations during filesystem access. To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked. In case two view objects are aliased, this operation results in a~\emph{no-op}.

Consequently, the simple example does not explicitly specify a memory space where the data lives in. However, Kokkos ensures that the defaults for parallel execution and data allocations match.

@@

While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources. Resources include example applications for individual Kokkos features, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation. Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}. The tutorial is intended for students with minimal or even no prior knowledge of parallel programming. It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises that build upon each other and gradually introduce new concepts. The programming guide is available on-line~\cite{KOKKOS_WIKI}.

@@

    Kokkos::parallel_for("outer", 
      TeamPolicy<>(N,AUTO),
      [=](const member_type &team_handle) {
        const int e = team_handle.league_rank();
        Kokkos::parallel_reduce( 
          TeamThreadRange(team_handle, M),
          [=](const int & i, double & y_tmp) {
            y_tmp += A(e, i) * x(i);
          }, 
          y(e));
    }); 
    Kokkos::fence();
    Kokkos::deep_copy(y_h,y);
    //output_result_on_host(y_h);
  }
  Kokkos::finalize();
}
\end{minted}
\end{small}
\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop using a ~\emph{TeamPolicy} and a nested parallel reduction. }
\label{fig:KokkosExample}
\end{figure}

>>>>

If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation. In real applications, a host-accessible allocation is necessary to allow I/O operations during filesystem access. To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked. In case two view objects are aliased, this operation results in a~\emph{no-op}. Consequently, the simple example does not explicitly specify a memory space where the data lives in. However, Kokkos ensures that the defaults for parallel execution and data allocations match.

@@

While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources. Resources include example applications for each individual Kokkos feature, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation. Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}. The tutorial is intended for students with previous knowledge of one parallel programming model and basic knowledge of C++. It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises that build upon each other and gradually introduce new concepts. The programming guide is available on-line~\cite{KOKKOS_WIKI}. In chapter~\ref{chap:teaching} we provide further details on best practices obtained from teaching Kokkos as well as list feedback received from lecture attendees.

@@

    Kokkos::parallel_for("outer", 
      TeamPolicy<>(N,AUTO),
      [=](const member_type &team_handle) {
        const int e = team_handle.league_rank();
        Kokkos::parallel_reduce( 
          TeamThreadRange(team_handle, M),
          [=](const int & i, double & y_tmp) {
            y_tmp += A(e, i) * x(i);
          }, y(e));
    }); 
    Kokkos::fence();
    Kokkos::deep_copy(y_h,y); //copy to host memory space    
  }
  Kokkos::finalize();
}
\end{minted}
\end{small}
\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop using a ~\emph{TeamPolicy} and a nested parallel reduction. }
\label{fig:KokkosExample}
\end{figure}

### 2020-08-15 src/6.Backend.tex
Updates for submission to EduHPC at SC20

\section{Back-end Support}\label{chap:backend}

In this section we peek inside the Kokkos library and show how a generic parallel for loop is mapped to the OpenMP backend programming model. As a template metaprogramming library, Kokkos makes use of partial specialization. In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes. Such classes correspond to a particular back-end where for each back-end, concepts with partial specialization must be provided. Today, Kokkos supports the CUDA, OpenMP, HIP as well as one-sided MPI, SHMEM and NVSHMEM back-ends. For each new back-end, partially implemented classes that implement necessary concepts must be provided.

A simplified version of a partially specialized class for a parallel for loop implementation is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}. The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function. That function is then responsible to implement the Kokkos parallel pattern. In this back-end, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled. 
\begin{figure}
\begin{small}
\begin{minted}[fontsize=\small]{c++}
template <class FunctorType>
class ParallelFor<FunctorType,
              RangePolicy<OpenMP>> 
{
  const FunctorType functor;
  const RangePolicy<OpenMP> policy; 
public:
  void execute() const {
    #pragma omp parallel for
    for(int i=policy.begin();
            i<policy.end();
            ++i)
      functor(i);
  }
};
template <class FunctorType,class PolicyType>
void parallel_for(std::string label, 
    const PolicyType& policy,
    const FunctorType& functor) {
  // Call profiling tool hooks with label
  ParallelFor<FunctorType,PolicyType> 
    pf{functor,policy};
  pf.execute();
}
\end{minted}
\end{small}
\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programming model. It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space. Also, it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
\label{fig:KokkosExampleOMPBackEnd}
\end{figure}

>>>>

\section{Backend Support}\label{chap:backend}

In this section, we peek inside the Kokkos library and show how a generic parallel for loop is mapped to the OpenMP backend programming model. As a template metaprogramming library, Kokkos makes use of partial specialization. In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} function calls, as shown in Figure~\ref{fig:KokkosExample}, are mapped to partially specialized classes. Such classes correspond to a particular backend where for each backend, concepts with partial specialization must be provided. Today, Kokkos supports the CUDA, OpenMP, HIP as well as one-sided MPI, SHMEM and NVSHMEM backends. For each new backend, partially implemented classes that implement necessary concepts must be provided.

A simplified version of a partially specialized class for a parallel for-loop implementation in the OpenMP backend is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}. The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function. That function is then responsible to implement the Kokkos parallel pattern. In this backend, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled. Figure~\ref{fig:stack} gives an architectural overview of Kokkos and lists important abstractions as well as the currently supported backends.
\begin{figure}
\begin{small}
\begin{minted}[fontsize=\small]{c++}
template <class FunctorType>
class ParallelFor<FunctorType,
              RangePolicy<OpenMP>> {
  const FunctorType functor;
  const RangePolicy<OpenMP> policy; 
public:
  void execute() const {
    #pragma omp parallel for
    for(int i=policy.begin();
            i<policy.end();
            ++i)
      functor(i);
  }
};
template <class FunctorType,class PolicyType>
void parallel_for(std::string label, 
    const PolicyType& policy,
    const FunctorType& functor) {
  // Call profiling tool hooks with label
  ParallelFor<FunctorType,PolicyType> 
    pf{functor,policy};
  pf.execute();
}
\end{minted}
\end{small}
\caption{This code shows the backend implementation of a parallel loop in the Kokkos programming model. It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space. Also, it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
\label{fig:KokkosExampleOMPBackEnd}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.42\textwidth]{img/Stack.png}}
\caption{Overview of the Kokkos programming model shows important abstractions as well as the inclusion of backend programming models used by the runtime to map higher level concepts to concrete hardware architectures.}
\label{fig:stack}
\end{figure}

### 2020-08-15 src/8.Related.tex
Updates for submission to EduHPC at SC20

C++ increasingly supports parallel constructs. It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers. However, at the time of publication, C++ does not support accelerators or many of the complexities of parallel computing related to accelerator programming. In courses that do not target accelerators, this can be a relevant option.

Over the past decade the Kokkos programming model has contributed to a number of current and near-future ISO-C++ features, including \emph{atomic\_ref}\cite{wg21_p0019}, \emph{mdspan}\cite{MDSPAN}, and executors.  In particular, the design of C++ executors, which are similar to Kokkos execution spaces, were influenced by this programming model. As the demand to extract performance from increasingly deep and increasingly asynchronous software stacks across a wide variety of domains increases, many C++ experts expect executors to become a central abstraction in any performance-oriented software stack.

RAJA~\cite{RAJA} is a C++ abstraction library with similarities to Kokkos. The important difference however is that Kokkos is descriptive, while RAJA is prescriptive. In Kokkos, the programming models determine how an application is mapped to the underlying hardware. RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures. RAJA is a viable option for class education in which the educator would like to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.

DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel architectures. In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns. Unlike Kokkos, it targets Intel architectures only and is not vendor-agnostic. A course targeting Intel architectures, DPC++ could be a valid option.

>>>>

C++ increasingly supports parallel constructs. It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers. However, at the time of publication, C++ does not support accelerators or many of the complexities of parallel computing related to accelerator programming. In courses that do not target accelerators, this can be a relevant option. Over the past decade the Kokkos programming model has contributed to a number of current and near-future ISO-C++ features, including \emph{atomic\_ref}\cite{wg21_p0019}, \emph{mdspan}\cite{MDSPAN}, and executors.  In particular, the design of C++ executors, which are similar to Kokkos execution spaces, were influenced by this programming model. As the demand to extract performance from increasingly deep and increasingly asynchronous software stacks across a wide variety of domains increases, many C++ experts expect executors to become a central abstraction in any performance-oriented software stack. 
RAJA~\cite{Raja} is a C++ abstraction library with similarities to Kokkos. The important difference however is that Kokkos is descriptive, while RAJA is prescriptive. In Kokkos, the programming model determines how an application is mapped to the underlying hardware. RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures. RAJA is a viable option for class education in which the educator would like to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel hardware architectures. In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns. Unlike Kokkos however, it targets Intel architectures only and is not vendor-agnostic. A course targeting Intel architectures, DPC++ could be a valid option.
